\chapter{Symbolic Program Execution}
In the last chapter we have seen how to prove the correctness of a recursive function via
\blue{computational induction}.  If a function is implemented via loops instead of recursion, then the method
of computational induction is not applicable.  Therefore, this section introduces the method of
\blue{symbolic execution}. \index{symbolic execution}  Using this method it is possible to verify the
correctness of programs that are implemented in an iterative fashion using loops. 
We will introduce this method via a simple example.  Consider the program shown in Figure
\ref{fig:power-iterative-annotated.stlx}. 

\begin{figure}[!h]
\centering
\begin{Verbatim}[ frame         = lines, 
                  framesep      = 0.3cm, 
                  labelposition = bottomline,
                  numbers       = left,
                  numbersep     = -0.2cm,
                  xleftmargin   = 1.3cm,
                  xrightmargin  = 1.3cm,
                  codes         = {\catcode`_=8\catcode`$=3},
                  commandchars  = \\\{\},
                ]
    def power(x$_1$, y$_1$):
        r$_1$ = 1
        while y$_n$ > 0:
            if y$_n$ % 2 == 1:
                r$_{n+1}$ = r$_n$ * x$_n$
            x$_{n+1}$ = x$_n$ * x$_n$
            y$_{n+1}$ = y$_n$ \symbol{92} 2
        return r$_N$
\end{Verbatim}
\vspace*{-0.3cm}
\caption{An annotated programm to compute powers.}
\label{fig:power-iterative-annotated.stlx}
\end{figure} % $

The main difference between a mathematical formula and a program is that in a formula all
occurrences of a variable refer to the same value.   This is different in a program because the
variables change their values dynamically.  In order to deal with this property of program variables
we have to be able to distinguish the different occurrences of a variable.  To this end,  we 
\blue{index} the program variables. 
When doing this we have to be aware of the fact that the same occurrence of a program variable can
still denote different values if the variable occurs inside  a loop.  In this case we have to index
the variables in a way that the index includes a counter that counts the number of loop iterations.
For concreteness, consider the  program shown in 
Figure \ref{fig:power-iterative-annotated.stlx}.  
Here, in line 5 the variable \mytt{r} has the index $n$ on the right side of the assignment,
while it has the index $r_{n+1}$ on the left side of the assignment in line 5.  The index $n$ denotes 
the number of times that the test \mytt{y$_n$ > 0} of the \mytt{while} loop has been executed.
After the loop finishes in line 10 the variable $\mytt{r}$ is indexed as
$\mytt{r}_N$, where $N$ denotes the total number of times that the test \mytt{y > 0} has been executed.
We show the correctness of the given program next.  Let us define
\\[0.2cm]
\hspace*{1.3cm}
$ a := x_1, \quad b := y_1$.
\\[0.2cm]
We will show that the \mytt{while} loop satisfies the invariant
\begin{equation}
  \label{eq:powerInv}
   r_n \cdot x_n^{y_n} = a^b.
\end{equation}
This claim is proven by induction on the number of loop iterations.
\begin{enumerate}
\item[B.C.:] $n=1$.

            Since we have $r_1 = 1$, $x_1 = a$, and $y_1 = b$ we have 
            \\[0.2cm]
            \hspace*{1.3cm}
            $r_n \cdot x_n^{y_n} = r_1 \cdot x_1^{y_1} = 1 \cdot a^{b} = a^b$.
\item[I.S.:] $n \mapsto n + 1$.

            We proof proceeds by a case distinction with respect to the expression $y \mmod 2$:
            \begin{enumerate}
            \item $y_n \mmod 2 = 1$.

                  Then we have $y_{n} = 2 \cdot (y_n\dv 2) + 1$ and
                  $r_{n+1} = r_n \cdot x_n$.  Hence
                  \begin{eqnarray*}
                      &   & r_{n+1} \cdot x_{n+1}^{y_{n+1}} \\[0.2cm] 
                      & = & (r_{n} \cdot x_n) \cdot (x_{n} \cdot x_{n})^{y_{n}\dvv 2} \\[0.2cm] 
                      & = & r_{n} \cdot x_n^{2 \cdot (y_{n}\dvv 2) + 1} \\[0.2cm] 
                      & = & r_{n} \cdot x_n^{y_n} \\
                      & \stackrel{i.h.}{=} & a^{b} 
                  \end{eqnarray*}

            \item $y_n \mmod 2 = 0$.

                  Then we have $y_{n} = 2 \cdot (y_n\dv 2)$ and $r_{n+1} = r_n$.
                  Therefore
                  \begin{eqnarray*}
                      &   & r_{n+1} \cdot x_{n+1}^{y_{n+1}} \\[0.2cm] 
                      & = & r_{n} \cdot (x_{n} \cdot x_{n})^{y_{n}\dvv 2} \\[0.2cm] 
                      & = & r_{n} \cdot x_n^{2 \cdot (y_{n} \dvv 2)} \\[0.2cm] 
                      & = & r_{n} \cdot x_n^{y_n} \\
                      & \stackrel{i.h.}{=} & a^{b} 
                  \end{eqnarray*}
            \end{enumerate}
\end{enumerate}
This shows the validity of equation (\ref{eq:powerInv}).   If the \mytt{while} loop
terminates, we must have $y_N = 0$.  If $n=N$, then equation (\ref{eq:powerInv}) yields:
\\[0.2cm]
\hspace*{1.3cm}
$$
\begin{array}[t]{cl}
                 & r_N \cdot x_N^{y_N} = a^b \\
\Leftrightarrow  & r_N \cdot x_N^{0}  = a^b \\
\Leftrightarrow  & r_N \cdot 1       = a^b \\
\Leftrightarrow  & r_N               = a^b
\end{array}
$$
\\[0.2cm]
This shows $r_N = a^b$ and since it is obvious that the \mytt{while} loop terminates, we have proven that
$\mytt{power}(a,b) =a^b$ holds. \qed

\exercise
Use the method of symbolic program execution to prove the correctness of the implementation of the
Euclidean algorithm that is shown in Figure \ref{fig:gcd.stlx} on page \pageref{fig:gcd.stlx}.  During the proof
you should make use of the fact that for all positive natural numbers $x$ and $y$ the equation
\\[0.2cm]
\hspace*{1.3cm}
$\mytt{gcd}(x, y) = \mytt{gcd}(x \,\mytt{\%}\, y, y)$
\\[0.2cm]
is valid.  

\begin{figure}[!ht]
\centering
\begin{Verbatim}[ frame         = lines, 
                  framesep      = 0.3cm, 
                  firstnumber   = 1,
                  labelposition = bottomline,
                  numbers       = left,
                  numbersep     = -0.2cm,
                  xleftmargin   = 0.8cm,
                  xrightmargin  = 0.8cm,
                ]
    def gcd(x, y):
        while y != 0:
            x, y = y, x % y
        return x
\end{Verbatim}
\vspace*{-0.3cm}
\caption{The Euclidean algorithm.}
\label{fig:gcd.stlx}
\end{figure}
\pagebreak

\section{Check Your Understanding}
If you were able solve the previous exercise, then you should have mastered the concepts
introduced in this chapter.


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "algorithms"
%%% End:
